Governing through patronage: the bargain for education in decentralized Brazil

### 2020-10-15 paper/paper.tex
redraft intro

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public sresources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}. The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.

Understanding how these political decisions cascade down the bureaucracy remains a challenge. Public services the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom \citep{}.

How can we intervene in this chain to improve the quality of public services? Previous studies have explored how politicians' incentives can improve the quality of public goods provision. Some have focused on the direct link between the political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}. Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce positive changes \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate. Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.

To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

>>>>

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public resources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}. The World Bank refers to this phenomenon as ``learning poverty'': stunted learning even when access to schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} There is growing recognition of the role bureaucratic institutions play in shaping these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}, but the relationship between politicians, bureaucrats and the quality of public services is less clear.

In particular, there is widespread debate whether political actors' discretion over bureaucratic institutions can have positive or negative consequences for public services. On the one hand, a large literature on state capacity has argued that political interference has detrimental consequences for state capacity and public goods provision \citep{evans_bureaucracy_1999,kohli_state-directed_2004, akhtari_political_2015}. On the other, recent studies have highlighted how greater political control can have a positive effect, as increased political oversight and credit claiming can lead to better developmental outcomes \citep{gulzar_politicians_2017, raffler_does_2016}. This dichotomy between politicians and bureaucrats frames our understanding of bureaucratic quality \citep{grindle_jobs_2012}, but unpacking this relationship requires taking a step in and out, recognizing the multiplicity of actors and institutions that frame the political administration of public services \citep{centeno_unpacking_2017}.

In much of the developing world, political actors do not govern or manage alone. Particularly in developing democracies, the political landscape is composed of multiple elected officials and power-sharing arrangements which ultimately shape the political nexus between elected officials and bureaucratic institutions \citep{francois_how_2015}. In the context of a separation of powers, the need for alignment between executive and legislative actors can give rise to political competition and the need to form coalitions that allow executive leaders to govern and implement their policy agenda \citep{cameron_models_2004, gottlieb_countervailing_2019}. Where programmatic commitments are weak, this coalition-formation -- known as presidential coalitionism --  manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}. Appointments into education, due to their high social prestige and potential for electoral mobilization \citep{oliveros_making_2016}, can be particularly valuable.

These patronage transactions between political elites have important consequences for bureaucratic institutions. As noted by \citet{evans_embedded_1995}, one feature of high-performing bureaucracies is their cohesion: an ability to retain and select personnel without being exposed to cycles of disruption. Patronage appointment

To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
